---
title: "Creating all processed data in the rnhanesdata package"
author: "Andrew Leroux"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


In this vignette we walk through the creation of the 8 primary processed datasets contained in the **rnhanesdata** pacakge. These 
datasets can be organized into four categories: 

- Activity count data (accelerometer)
- Wear/non-wear flags associated (derived from the activity count data)
- NHANES demographic, survey design, lifestyle, and comorbiditiy variables
- Mortality data linked to NHANES


## Processing the accelerometry data and creating wear/non-wear flags


```{r, process_accel_flags, eval=FALSE}
accel_ls <- process_accel(names_accel_xpt = c("PAXRAW_C","PAXRAW_D"),
                          local=FALSE, deleteraw=TRUE, write=TRUE,
                          urls=c("https://wwwn.cdc.gov/Nchs/Nhanes/2003-2004/PAXRAW_C.ZIP",
                                 "https://wwwn.cdc.gov/Nchs/Nhanes/2005-2006/PAXRAW_D.ZIP")
                          )
flags_ls <- process_flags(accel_ls,write=TRUE,localpath="~/Downloads/")
```

## Processing the mortality data 


## Processing the demographic, survey design, lifestyle, and comorbiditiy variables






<!-- Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format: -->

<!-- - Never uses retina figures -->
<!-- - Has a smaller default figure size -->
<!-- - Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style -->

<!-- ## Vignette Info -->

<!-- Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette. -->

<!-- ## Styles -->

<!-- The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         css: mystyles.css -->

<!-- ## Figures -->

<!-- The figure sizes have been customised so that you can easily put two images side-by-side. -->

<!-- ```{r, fig.show='hold'} -->
<!-- plot(1:10) -->
<!-- plot(10:1) -->
<!-- ``` -->

<!-- You can enable figure captions by `fig_caption: yes` in YAML: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         fig_caption: yes -->

<!-- Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**. -->

<!-- ## More Examples -->

<!-- You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`. -->

<!-- ```{r, echo=FALSE, results='asis'} -->
<!-- knitr::kable(head(mtcars, 10)) -->
<!-- ``` -->

<!-- Also a quote using `>`: -->

<!-- > "He who gives up [code] safety for [code] speed deserves neither." -->
<!-- ([via](https://twitter.com/hadleywickham/status/504368538874703872)) -->
